Dissecting the causal question underlying the association between cancer and ADRD 

### 2021-05-05 04_reports/manuscript.Rmd
updated manuscript and dags

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [pdf_document]
editor_options: 
  chunk_output_type: console
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(kableExtra)
# source(here::here("02_R", "03_descriptives.R"))
# source(here::here("02_R", "05_plots.R"))
# source(here::here("02_R", "05b_plot_causes_death.R"))
# table_results <- rio::import(here::here("02_R", "table_results.csv"))
# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
```

@@

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

@@

\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

However, if the drug has any side-effects we have to decide if either we want to estimate a total effect (that is, the effect on ADRD even if partly mediated by the effect on death) or a direct effect (that is, the effect directly on ADRD not mediated by death) (ref). We face this challenge in our observational data setting, since we are measuring cancer diagnosis as the proxy for Pin1 over-expression, and cancer is a leading cause of death29. We can visualize this data feature in the causal diagram in Figure 5. In this causal diagram we include $D_{t+1}$ as a representation of death over follow-up, an arrow between $P_0$ and $D_{t+1}$ since $P_0$ may increase the mortality risk. We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C_2$ represent the shared causes of both events.

@@

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [html_document]
editor_options: 
  chunk_output_type: console
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(kableExtra)
# source(here::here("02_R", "03_descriptives.R"))
# source(here::here("02_R", "05_plots.R"))
# source(here::here("02_R", "05b_plot_causes_death.R"))
# table_results <- rio::import(here::here("02_R", "table_results.csv"))
# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
```

@@

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=45,in=135, looseness=0.2] (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

@@

\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm,](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

However, if the drug has any side-effects we have to decide if either we want to estimate a total effect (that is, the effect on ADRD even if partly mediated by the effect on death) or a direct effect (that is, the effect directly on ADRD not mediated by death) (ref). We face this challenge in our observational data setting, since we are measuring cancer diagnosis as the proxy for Pin1 over-expression, and cancer is a leading cause of death29. We can visualize this data feature in the causal diagram in Figure 5. In this causal diagram we include $D_{t+1}$ as a representation of death over follow-up, an arrow between $P_0$ and $D_{t+1}$ since $P_0$ may increase the mortality risk. We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C$ represent the shared causes of both events.

@@

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

@@

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```
